\input{../../preamble}
\begin{document}
\titleheader{Empty Promises}
I would like to dedicate this final discussion to vacuous statements - statements that are true \textit{by default} - or by lack of counterexamples.

Let me show you.

Suppose you have an empty bucket. Consider the statement \texttt{all balls in this bucket are blue}. Is this statement true or false?

Well, can you produce a ball from the bucket that \textit{isn't} blue? No, of course not. Thus the statement must be true.

Now you could argue that one cannot produce a blue ball either. But get this, my statement was \texttt{if a ball is in this bucket, it is blue}. You \textit{cannot} disprove it, because doing so would require you to produce a counterexample from the bucket. And as we know, the bucket is empty. Hence there are no counterexamples.

\section*{A Little bit More}
Vacuous statements have the form $\forall \space r, P(r) \implies Q(r)$. The difference is that $r$ either comes from an empty collection, or for every $r$, $\neg P(r)$ holds. (Recall the equivalent of an implication).

Have some more examples.
\begin{enumerate}
    \item Every even injective function from $\mathbb R$ to itself is continuous
    \item Every skew-symmetric nonzero $1\times 1$ matrix is invertible.
    \item Every even integer of the form $24k+13$ is a perfect square
\end{enumerate}
and so on. This is important because sometimes, it could be the case that a forall statement is vacuously true, but you use it the wrong way.

I really don't have any examples for this other than the footnote in PW-13.

\section*{Conclusion}
That concludes my little course on proofwriting and logic. I hope it was useful to you, and that you learned something. 

If you encounter questions, typos, or anything else, please feel free to reach out. Cya.
\end{document}